Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-06-09 response-to-reviews.Rmd

-Table 2, gb unit, mmol m-2 s-1 ?

*Nidhi*

-I found the x axes labels in Fig 4.a difficult to understand, there is maybe a more intuitive way to write them. Note that PET is not defined. °C for the units.

**We have fixed the axes labels to list units as °C and updated the caption to define PET and better explain the x-axis labels.**

@@

l.187: 'but the elevated CO2 quickly dissipates during the day such that differences are negligible for most of the day when photosynthesis is occurring'. Please specify 'CO2 concentration differences' where it says differences, to increase clarity.

**Revised as suggested.**

l.200 'upper canopy leaves have higher 𝑔𝑏 and associated potential for water and heat loss' compared to what??? Please clarify.

**Done. The sentence now reads: "The implication is that upper canopy leaves have higher *g~b~* and associated potential for water and heat loss compared to understory leaves (sections 2, 4.1)."**

### 2022-06-09 response-to-reviews.Rmd

**(2) modified the abstract sentence to read, "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**

**(3) changed statement summarizing @rollinsonClimateSensitivityUnderstory2020 to a more generic statement: "Indeed, there are cases where tree-ring records show more negative or less positive $T_{air}$-responses of understory trees compared to canopy dominants [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020]."**

**(4) changed the statement, "...and tree ring evidence (Fig. 4) suggest that understory trees may respond more negatively to hot, wet conditions." to read "...and some tree ring evidence [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020] raise the possibility that understory trees may be at a relative disadvantage under hot, humid conditions."**

**(5) revised discussion in the "Warming" section to read: "In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a]. However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010]. Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions."**

**(6) in the conclusions section, removed the statement  "...[understory trees] might be more vulnerable to chronic warming stress under relatively mesic conditions [@rollinsonClimateSensitivityUnderstory2020]"**

@@

l.579 'implies that high 𝑇𝑎𝑖𝑟 should decrease 𝑔𝑠 of upper canopy leaves more than understory leaves'. The authors here are ignoring what they’ve said previously, that high Tair is associated with high VPD (l. 199-200), and that VPD is conclusively a much stronger driver of gs than Tleaf (Leuning 1995, Grossiord). I think the sentence should be corrected and include VPD.  I'm not yet aware of a good study that has decoupled VPD and Tleaf to prove that Tleaf intrinsically causes reduction in gs. So the authors cannot conclude this.

l.629 'we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'. Other than that it varies according to the prevailing temperature difference? I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this.

l.806-807 'trees in the understory might be more negatively affected by chronic stress from warming 𝑇𝑎𝑖𝑟'. This is based on such very thin evidence, almost no evidence. Nor do I think that the  authors are appropriately highlighting a lack of knowledge very well by their speculation (which is not based on data). They are inferring what would happen with climate warming on the basis of historical evidence, highly indirect in Fig. 4B which only shows modelled results, and when there are highly overlapping CI's. I hardly find it a basis for a strong statement landing in the abstract, especially when it is contradicted by other statements throughout the review.

**We have revised the second half of the paragraph containing this statement as follows: "In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a]. However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010]. Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions."**

**We have modified the abstract sentence to read, "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**

### 2022-06-09 vertical-thermal-review.Rmd
update Fig. 4
#119
edit discussion of small trees
#120

1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA

@@

![**Figure `r fig_treerings`. Examples of tree-ring evidence for greater interannual temperature sensitivity of annual growth for large than for small trees.** Across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high maximum temperature (*T~max~*) or potential evapotranspiration (PET) than did small trees of the same species. Independent variables are the most influential temperature-related variable at the site (*T~max~* or PET over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively). Colored lines represent responses  of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/pre_temp_groups_dbh_interactions_for_NIDHI.png)

@@

It remains uncertain how responses to warming will vary across vertical gradients under mesic conditions. 
Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
As growing season *T~air~* increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and *T~crit~* [@osullivanThermalLimitsLeaf2017], *T~leaf~* of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress]. 
In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~* (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher *T~air~*.
Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
<!--Moreover, increased nighttime *T~air~* may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [@rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming *T~air~*, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].

@@

Across vertical gradients, directional trends in the biophysical environment and leaf traits are the rule, driving variation in the physiology and ecology that have these as their underpinnings (Fig. `r fig_schematic`).
However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. 
While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions [@rollinsonClimateSensitivityUnderstory2020].

>>>>

1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA

@@

![**Figure `r fig_treerings`. Examples of tree-ring evidence for greater interannual temperature sensitivity of annual growth for large than for small trees.** Across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high maximum temperature (*T~max~*) or potential evapotranspiration (PET) than did small trees of the same species. Independent variables are the most influential temperature-related variable at the site (*T~max~* or PET over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively). Colored lines represent responses  of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/tree_ring_figure.png)

@@

It remains uncertain how responses to warming will vary across vertical gradients under mesic conditions. 
Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
As growing season *T~air~* increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and *T~crit~* [@osullivanThermalLimitsLeaf2017], *T~leaf~* of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress]. 
In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
<!--Moreover, increased nighttime *T~air~* may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010].
Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions.

@@

Across vertical gradients, directional trends in the biophysical environment and leaf traits are the rule, driving variation in the physiology and ecology that have these as their underpinnings (Fig. `r fig_schematic`).
However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. 
While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees.
